---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# effectSizeMeasures

<!-- badges: start -->
![R-CMD-check](https://github.com/neokok/effectSizeMeasures/actions/workflows/R-CMD-check.yaml/badge.svg)
[![Codecov test coverage](https://codecov.io/gh/neokok/effectSizeMeasures/graph/badge.svg)](https://app.codecov.io/gh/neokok/effectSizeMeasures)

<!-- badges: end -->

The goal of effectSizeMeasures is to provides tools to calculate different effect size metrics for statistical analysis. It allows for efficient calculations of five different effect size measures: Pearson's correlation coefficient, coefficient of determination, odds ratio, risk ratio (relative risk), and Cohen's d. 

## Installation

You can install the development version of effectSizeMeasures from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("neokok/effectSizeMeasures")
```

## Example

These are basic examples which show you how to solve utilize the functions:
```{r setup, include = FALSE}
devtools::load_all(".")
```

```{r example}
library(effectSizeMeasures)
```

Example use of calculating Pearson's correlation coefficient:
```{r example 1}
pearson_corr(c(1, 2, 3, 4), c(4, 6, 7, 4))
```

Example use of calculating coefficient of determination:
```{r example 2}
coef_determination(c(1, 2, 3, 4), c(4, 6, 7, 4))
```

Example use of calculating odds ratio with the following table:

|         | Diseased | Non-Diseased |
|---------|----------|--------------|
| Exposed  | 20          | 380          |
| Non-Exposed | 6          | 594          |

```{r example 3}
odds_ratio(matrix(c(20, 6, 380, 594), nrow = 2, ncol = 2))
```

Example use of calculating risk ratio with the following table:

|         | Diseased | Non-Diseased |
|---------|----------|--------------|
| Exposed  | 20          | 380          |
| Non-Exposed | 6          | 594          |

```{r example 4}
risk_ratio(matrix(c(20, 6, 380, 594), nrow = 2, ncol = 2))
```

Example use of calculating Cohen's d:
```{r example 5}
cohens_d(c(1, 2, 3, 4), c(4, 6, 7, 4), TRUE)
```
